---
title: "04 Apply family functions - task 2"
output: html_document
---

<!--
# ASSIGNMENT CONFIG
init_cell: false
export_cell: false
generate:
    show_stdout: true
    show_hidden: false
-->


<!--
# BEGIN QUESTION
name: q2
manual: false
points:
    - 3
    - 1
-->

**Question 2:** Create a function `CheckSize()` that takes as input any list and returns:
* string `List is empty` if the input is an empty list
if not, a list (of the same length as an input list), each element of which is either:
* string `Object is too small` if a length of a given element is smaller than 2
* string `Object is too large` if a length of a given element is greater than 5
* element itself otherwise

Let's see some examples:
`list()` should return `List is empty`
`list(4, 1:5, matrix(1:4, 2, 2), 'I like R', matrix(1:16, 4, 4), c('L', 'A', 'Z', 'A', 'R', 'S', 'K', 'I'))` should return
`[[1]]`
`[1] "Object is too small"`

`[[2]]`
`[1] 1 2 3 4 5`

`[[3]]`
`     [,1] [,2]`
`[1,]    1    3`
`[2,]    2    4`

`[[4]]`
`[1] "Object is too small"`

`[[5]]`
`[1] "Object is too large"`

`[[6]]`
`[1] "Object is too large"`

<!-- # BEGIN SOLUTION -->

```{r}
# BEGIN SOLUTION NO PROMPT
AuxFun <- function(elem){
  if (length(elem) < 2) {'Object is too small'}
  else if (length(elem) < 6) {elem}
  else {'Object is too large'}
}

CheckSize <- function(my_list){
  if (length(my_list) == 0) return('List is empty')
  return(lapply(my_list, AuxFun))
}
# END SOLUTION
# YOUR CODE HERE
# END PROMPT
```

<!-- # END SOLUTION -->

<!-- # BEGIN TESTS -->


```{r}
# HIDDEN
list1 <- list(22, 1:3, matrix(1:64, 8, 8), 'I like R', matrix(1:3, 3, 1), c('L', 'A', 'Z', 'A', 'R', 'S', 'K', 'I'))
testthat::expect_true(all(CheckSize(list1)[[1]] == 'Object is too small'))
testthat::expect_true(all(CheckSize(list1)[[2]] == 1:3))
testthat::expect_true(all(CheckSize(list1)[[3]] == 'Object is too large'))
testthat::expect_true(all(CheckSize(list1)[[4]] == 'Object is too small'))
testthat::expect_true(all(CheckSize(list1)[[5]] == matrix(1:3, 3, 1)))
testthat::expect_true(all(CheckSize(list1)[[6]] == 'Object is too large'))
testthat::expect_true(typeof(CheckSize(list1)[[1]]) == 'character')
testthat::expect_true(typeof(CheckSize(list1)[[2]]) == 'integer')
testthat::expect_true(typeof(CheckSize(list1)[[3]]) == 'character')
testthat::expect_true(typeof(CheckSize(list1)[[4]]) == 'character')
testthat::expect_true(typeof(CheckSize(list1)[[5]]) == 'integer')
testthat::expect_true(typeof(CheckSize(list1)[[6]]) == 'character')
```

```{r}
# HIDDEN
list2 <- list()
testthat::expect_true(CheckSize(list2) == 'List is empty')
testthat::expect_true(typeof(CheckSize(list2)) == 'character')
```
<!-- # END TESTS -->

<!-- # END QUESTION -->