---
title: "04 Apply family functions - task 3"
output: html_document
---

<!--
# ASSIGNMENT CONFIG
init_cell: false
export_cell: false
generate:
    show_stdout: true
    show_hidden: false
-->


<!--
# BEGIN QUESTION
name: q3
manual: false
points:
    - 4
-->

**Question 3:** Create a function `SepNumWithLet()` that receives as an input 2 arguments, that are numeric vectors. Assume that these vectors are of the same length `n`, where `0 < n <= 26`. (You don't need to hardcode these constraints, just assume that the input arguments satisfy these conditions).
This function should return the character vector that pastes two numeric vectors with the separator of the first `n` lower-case letters in the alphabet as follows: If the first input argument is `vec1 = c(vec11, vec12, ..., vec1n)` and the second one is `vec2 = c(vec21, vec22, ..., vec2n)`, then the output should be `"vec11avec21", "vec12bvec22", ..., "vec1nnvec2n"`.

Let's see some examples:
`SepNumWithLet(c(1, 2, 3), c(10, 20, 50))` should return `"1a10", "2b20", "3c50"`.
`SepNumWithLet(rep(3, 5), 5:9` should return `"3a5", "3b6", "3c7", "3d8", "3e9"`.

Hints:
* Use the multivariate version of apply `mapply()`.
* To extract letters you can use `letters` constant already defined in R.

<!-- # BEGIN SOLUTION -->

```{r}
# BEGIN SOLUTION NO PROMPT
SepNumWithLet <- function(vec1, vec2){
  len_vec <- length(vec1)
  return(mapply(paste, vec1, vec2, sep=letters[1:len_vec]))
}
# END SOLUTION
# YOUR CODE HERE
# END PROMPT
```

<!-- # END SOLUTION -->

<!-- # BEGIN TESTS -->


```{r}
# HIDDEN
testthat::expect_true(all(SepNumWithLet(c(3, 8), c(2, 4)) == c("3a2", "8b4")))
testthat::expect_true(all(SepNumWithLet(rep(5, 5), 1:5) == c("5a1", "5b2", "5c3", "5d4", "5e5")))
```

<!-- # END TESTS -->

<!-- # END QUESTION -->