---
title: "05 Data manipulation - task 3"
output: html_document
---

```{r}
library(dplyr)
```

**Question 3:** You have been given the `mtcars` dataset, which contains data on different car models and their characteristics. Your task is to use the `dplyr` package to:
* Group the data by the number of cylinders (`cyl`) and the transmission type (`am`).
* Calculate the average horsepower (`hp`) and the average miles per gallon (`mpg`) for each group. Name the variables `mean_hp` and `mean_mpg`, respectively.
* Select only the groups where the average horsepower (`mean_hp`) is greater than `150` and the average miles per gallon (`mean_mpg`) is less than `20`.
* Sort the groups by the average horsepower (`mean_hp`) in ascending order.

The resulting data frame (name it `output`) should have four columns: `cyl`, `am`, `mean_hp`, and `mean_mpg`.

Hints:
* `group_by()` and `summarize()` may be useful.
* You can use the pipe `%>%` operator.

```{r}
# YOUR CODE HERE
output <- mtcars %>%
  group_by(cyl, am) %>%
  summarize(mean_hp = mean(hp), mean_mpg = mean(mpg)) %>%
  filter(mean_hp > 150 & mean_mpg > 20) %>%
  arrange(mean_hp)
```
