```{r}
library(dplyr)
```

**Question 3:** You have been given the `mtcars` dataset, which contains data on different car models and their characteristics. Your task is to use the `dplyr` package to:
* Group the data by the number of cylinders (`cyl`) and the transmission type (`am`).
* Calculate the average horsepower (`hp`) and the average miles per gallon (`mpg`) for each group. Name the variables `mean_hp` and `mean_mpg`, respectively.
* Select only the groups where the average horsepower (`mean_hp`) is greater than `150` and the average miles per gallon (`mean_mpg`) is less than `20`.
* Sort the groups by the average horsepower (`mean_hp`) in ascending order.

The resulting data frame (name it `output`) should have four columns: `cyl`, `am`, `mean_hp`, and `mean_mpg`.

Hints:
* `group_by()` and `summarize()` may be useful.
* You can use the pipe `%>%` operator.
* Package dplyr is loaded in the previous chunk, so there is no need to repeat this action.

```{r}
# YOUR CODE HERE
```
